A Review of Mortality Models

### 2020-09-15 mortality-models/draft.tex
Spell check and other minor changes.

\documentclass{article}
\usepackage{amsmath, amssymb, amsthm, amsfonts}
\usepackage{url}
\numberwithin{equation}{section}
\DeclareMathOperator{\pr}{P}
\DeclareMathOperator{\E}{E}
\linespread{1.3}
\title{A Review of Mortality Models}
\author{To be filled}
\date{Sep-2020}
\begin{document}
\maketitle
\section{Introduction}\label{s0}
Although it is very difficult to predict the behavior of a member of a large
population one can often make statistical statements about the same population 
with much greater 
confidence. For example, it is impossible to predict when a particular 
radioactive atom will decay but one can make an accurate prediction of how 
many atoms will decay in a long enough time interval. Likewise, it is not
possible to predict the time of death of an individual although one can reliably
predict how many people from a large enough group will survive past a certain
date in future. Actuaries knew this statistical fact long before the 
mathematical foundations of the theory of probability were laid. The 
mathematical models they to describe how deaths occur in a given society are
called `mortality models'.

The death of an individual in a population or the decay of a radioactive atom
is an event happening to a member of the population. Likewise, the failure of
a machine, an occurrence of an event, the duration of a certain phase in an
economy are all different manifestations of the same phenomenon in which 
something happens that significantly alters the future of the member of the
population. Death terminates an individual's existence, radioactive decay
transforms the atom, failure renders a machine unusable and economies change 
after occurrences of certain cataclysmic events. Mortality of an individual
is akin to reliability of a machine. Life of an individual is similar to
the duration of a certain phase of business cycle. As a result, the 
theoretical tools used to study these outwardly different phenomena are similar 
to each other. 
Mortality analysis in actuarial science is called survival analysis is 
statistics, reliability theory in engineering, duration analysis in economics 
and event history analysis in sociology.

@@

However, the advancement in health science since the early years of the 
twentieth century, the relative peace in the western world since the second 
world war and abundance of food in the rich world resulted in a change in
historical mortality patterns. Pension funds started realizing that annuities
were being drawn for longer than expected periods. As the population in the
developed countries started graying the pension funds had the additional 
problem of a higher annuity obligations and a lower contribution from newer
workers. Pension funds now have a keen interest in understanding how mortality 
rates will evolve in the future. They now want to forecast mortality than just
produce models of the current mortality.

@@

Mortality models, being mathematical functions, produce a single curve for
a given set of its parameters. On the other hand, mortality forecasts are a
collection of curves or a family of probability distributions. Their output
is accompanied with appropriate confidence intervals to indicate the 
uncertainty in their reported values.

@@

We will sketch the historical development of mortality model, both 
deterministic and stochastic, starting from de Moivre's model in the next
section. We also define basic terms commonly used in actuarial science and 
demonstrate how to evaluate their expressions in simple cases. In section 3
we describe the Gompertz's model, the first realistic mortality model that 
was used for a very long time in the industry. The success of Gompertz model
lay in the number of successors it produced. Of these, we cover Makekam's 
model and its generalizations in section 4. In section 5 we mention a few
other mortality models used in European countries but whose significance today
lies mostly in their historical value. The $8$-parameter Heligman-Pollard model
was the first one that could be fitted to the entire mortality curve, starting
from infancy to very advanced years. We describe it in section 6. Starting
section 7 we turn our attention to stochastic models, starting with the
celebrated Lee-Carter model. An interesting class of models used across social
sciences is the Age-Period-Cohort models. We review a few of them relevant
to mortality in section 8. In section 9 we consider the Cairns-Blake-Dowd model
and finally we consider stochastic models with cohort effects in section 10. 
The models considered so far have looked at mortality irrespective of the
cause of death. However, in recent years, death rate due to certain causes
is falling while those due to others is rising. One can, therefore, forecast
death rates by cause and aggregate the results. This is the approach taken by
the cause-of-death models in section \ref{s11}. Several mortality models can
be written in the form $f(\mu) = L(\theta)$ where $L$ is a linear function of
the parameters $\theta$ of the model and $f$ is some function of the force
of mortality that permits us to have the right hand side as $L(\theta)$. The
statistically rigorous method to fit such models to mortality data is called
the generalized linear models. Section \ref{s12} discusses them. Not all
models can be brought in this format. Important examples being Lee-Carter and
Renshaw-Haberman. One has to take recourse to spline regression in these
cases. They are the topic of section \ref{s13}.

@@

Gompertz's model and its descendants considered in this section had simple
algebraic forms, each term of which could be interpreted in biological
terms. However, they were not always easy to fit on available mortality
data. The presence of exponentials also made them numerically unstable. 
Moreover, they covered only a limited number of sources of mortality.

@@

\section{Generalized linear models}\label{s12}
The ordinary least square linear regression assumes that the residuals are
normally distributed. The expected value of the response is a linear
combination of the predictors. Therefore a constant change in the predictor
results in a proportional change in the respose, a characteristic of a linear
relationship. However, often times this is not true. A generalized linear 
model expresses a certain function of the response as a linear model. That is,
instead of fitting a linear model between a response $y$ and the predictors $X$,
we git a linear model between a function $g(y)$ of the response and the
predictors. The function $g$ is called the link function. This method,
developed by Nelder and Wedderburn \cite{nelder1972generalized} proposes a 
general structure for several types of regression including the traditional
ordinary-least-squares, the logistic and the Poisson regressions.

If the probablity mass or density function of a random variable $Y$ can be
written as
\begin{equation}\label{s12e1}
f_Y(y, \theta; \phi) = 
\exp\left(\frac{y(\theta) - b(\theta)}{\alpha(\phi)} + c(y, \phi)\right)
\end{equation}
where $a, b, c$ are functions of their arguments and $\theta, \phi$ are
parameters then $Y$ is said to belong to an exponential family of distributions.The normal, binomial, Poisson, exponential and gamma distributions are all
members of the exponential family. A generalized linear model assumes that
the responses $Y_1, \ldots, Y_n$ are independent and identically distributed
with the distribution from the exponential family.

@@

If we write $B = e^{k_1}$ in equation \eqref{s3e3} describing the force of
mortality of Gompertz' law then we can take logarithm of both sides to get
\begin{equation}\label{e12e2}
\log\mu(x) = k_1 + kx.
\end{equation}
The right hand side of this equation is a linear function in $x$ while the left
hand side is a logarithm. This is precisely the form of generalized linear
model with logarithmic link function and a Poisson distrubution. If the errors
in Gompertz's law are modeled in terms of a binomial distribution then the
link function is a complimentary log-log function \cite{currie2013fitting} 
defined as
\begin{equation}\label{e12e3}
f(x) = \log(-\log(1 - x)),
\end{equation}
where $0 < x < 1$. Mortality laws expressed as sigmoid functions of the form
\begin{equation}\label{e12e4}
q_x = \frac{e^{\theta_0 + \theta_1 x}}{1 + e^{\theta_0 + \theta_1 x}},
\end{equation}
can be easily transformed to the form
\begin{equation}\label{e12e5}
\frac{q_x}{1 - q_x} = e^{\theta_0 + \theta_1 x}
\end{equation}
or, taking logarithm of both sides,
\begin{equation}\label{e12e6}
\log\frac{q_x}{1 - q_x} = \theta_0 + \theta_1 x.
\end{equation}
Once again we observe that the right hand side is a linear function of the
predictor. The function on the left hand is called a logit function and it is
a legitimate link function for a generalized linear model. In fact, this link
function is used to implement the standard logistic regression in applied
statistics. It is possible to show that the Age-Period-Cohort models and the
Cairns-Blake-Dowd models can also be expressed as generalized linear models. 
The algebra involved to demonstrate this fact is a bit involved and we refer
to the paper by Currie \cite{currie2013fitting} for the details.

@@

Splines were used to draw a smooth curve passing through a set of points. They
were initially used by draftsmen who represented the points by heavy leads and
the splines by flexible wooden strips. This idea was borrowed into numerical
analysis by I. J. Shoenberg \cite{schoenberg1969cardinal} in the late sixties
of the twentieth century. A set of polynomials was used to approximate a
non-linear function where each polynomial fitted a portion of the function
being approximated. It was possible to show that under certian conditions
\cite{schoenberg1969cardinal} a non-linear function could be expressed as a 
piecewise polynomial function. The polynomials were required to blend into each
other that the boundaries in a smooth way. That is, not only were their values
required to match but also their derivatives.

Splines can be implemented using the Cox-de Boor recurrence relation
\cite{de1978practical}
\begin{equation}\label{s13e1}
B_{i,0}(x) = \begin{cases}
1 & \text{ if } x \in [t_i, t_{i+1}] \\
0 & \text{ otherwise}
\end{cases}
\end{equation}
and
\begin{equation}\label{s13e2}
B_{i,j}(x) = \frac{x - t_i}{t_{i+j} - t_i}B_{i,j-1}(x) + 
\frac{t_{i+j+1}-x}{t_{i+j+1}-t_{i+1}}B_{i+1,j-1}(x).
\end{equation}
These splines are then joined at the knot points $\{t_i: i = 1, \ldots, n\}$.
Once the knots are given it is easy to find the splines. However, there is
little guidance on the number of knots. Too few of them lead to under-fitting
and too many of them lead to over-fitting. O'Sullivan \cite{o1986statistical}
suggested that one should use a relatively large number of knots and prevent
over-fitting by levying a penalty on the second derivatives of the splines.
Eilers and Marx \cite{eilers1996flexible} simplify O'Sullivan's suggestion
and apply it to regression on B-splines. They applied the penalty to the
difference between the coefficients of the adjacent splines. The optimal amount
of smoothing is determined by cross-validation and the Akaike Information 
Criterion (AIC). Eilers and Marx call splines with penalties as P-splines.
They demonstrate that P-splines are an extension of the generalized linear
models. P-spline transforms conserve statistical moments of the data and have
polynomial curve fits as limits. Currie and others \cite{currie2004smoothing}
show how to use P-splines to fit mortality forecasting models. If $x_a = (
x_{a,1},\ldots,x_{a,n_a})^T$ and $x_y=(x_{y,1},\ldots,x_{y,n_y})^T$ are the
vectors with ages and years, $\mathcal{B}_a = \{B_{a,1}, \ldots, B_{a, c_a}\}$
and $\mathcal{B}_y = \{B_{y,1},\ldots,B_{y, c_y}\}$ are the cubic B-slines of
dimensions $c_a$ and $c_y$ then
\begin{equation}\label{c14e1}
X = B_y \otimes B_a
\end{equation}
is the regression matrix for the P-spline model
\begin{equation}\label{c14e2}
M(x,t) = \sum_{i,j}\theta_{i,j}B_{i,j}^{ay}(x, t).
\end{equation}
Here $B_a =(B_{a,j}(x_{a,i}))$ is an $n_a \times c_a$ matrix and $B_y = 
(B_{y,j}(x_{y,i}))$ is an $n_y \times c_y$ matrix. The penalty matrix is given
as
\begin{equation}\label{c14e3}
P = \lambda_aI_{c_y} \otimes D_a^TD_a + \lambda_y D_y^TD_y\otimes I_{c_a},
\end{equation}
where $D_a$ is the difference matrix of the rows of M, $D_y$ is the difference
matrix on the columns of M, and the two $\lambda$s are the smoothing parameters.
The regression problem is then minimizing the likelihood function
\begin{equation}\label{c14e4}
\ell_p = \ell - \frac{1}{2} m^TPm,
\end{equation}
where $\ell$ is the default log likelihood function and $m$ is the vector of
the elements of $M$.

>>>>

\documentclass{article}
\usepackage{amsmath, amssymb, amsthm, amsfonts}
\usepackage{url}
\numberwithin{equation}{section}
\DeclareMathOperator{\pr}{P}
\DeclareMathOperator{\E}{E}
\linespread{1.3}
\title{A Review of Mortality Models}
\author{To be filled}
\date{Sep-2020}
\begin{document}
\maketitle
\section{Introduction}\label{s0}
Although it is very difficult to predict the behavior of a member of a large
population one can often make statistical statements about the same population 
with much greater 
confidence. For example, it is impossible to predict when a particular 
radioactive atom will decay but one can make an accurate prediction of how 
many atoms will decay in a long enough time interval. Likewise, it is not
possible to predict the time of death of an individual although one can reliably
predict how many people from a large enough group will survive past a certain
date in future. Actuaries knew this statistical fact long before the 
mathematical foundations of the theory of probability were laid. The 
mathematical models they use to describe how deaths occur in a given society are
called `mortality models'.

The death of an individual in a population or the decay of a radioactive atom
is an event happening to a member of the population. Likewise, the failure of
a machine, the extinction of a species, the duration of a certain phase in an
economy are all different manifestations of the same phenomenon in which 
something happens that significantly alters the future of the member of the
population. Death terminates an individual's existence, radioactive decay
transforms the atom, failure renders a machine unusable and economies change 
after occurrences of certain cataclysmic events. Mortality of an individual
is akin to reliability of a machine. Life of an individual is similar to
the duration of a certain phase of business cycle. As a result, the 
theoretical tools used to study these outwardly different phenomena are similar 
to each other. 
Mortality analysis in actuarial science is called survival analysis is 
statistics, reliability theory in engineering, duration analysis in economics 
and event history analysis in sociology.

@@

However, the advancement in health science since the early years of the 
twentieth century, the relative peace in the western world since the second 
world war and abundance of food in the rich world resulted in a change in
historical mortality patterns. Healthcare systems began to come under stress.
Pension funds started realizing that annuities
were being drawn for longer than expected periods. As the population in the
developed countries started graying the pension funds had the additional 
problem of a higher annuity obligations and a lower contribution from newer
workers. Pension funds now have a keen interest in understanding how mortality 
rates will evolve in the future. They now want to forecast mortality than just
produce models of the current mortality.

@@

Mortality models, being mathematical functions, produce a single curve for
a given set of its parameters. On the other hand, mortality forecasts are a
collection of curves or a family of probability distributions. Their output
is accompanied with appropriate confidence intervals to indicate the 
uncertainty in their reported values. Mortality models describe the mortality
behavior at a fixed time. They do not predict the changes in mortality over
long enough durations of the pension funds' interest. Mortality forecasts
allow us to extrapolate the mortality curves several decades into the future.

@@

We will sketch the historical development of mortality model, both 
deterministic and stochastic, starting from de Moivre's model in the next
section. We also define basic terms commonly used in actuarial science and 
demonstrate how to evaluate their expressions in simple cases. In section 3
we describe the Gompertz's model, the first realistic mortality model that 
was used for a very long time in the industry. The success of Gompertz model
lay in the number of successors it produced. Of these, we cover Makekam's 
model and its generalizations in section \ref{s3}. In section \ref{s4} we 
mention a few
other mortality models used in European countries but whose significance today
lies mostly in their historical value. The $8$-parameter Heligman-Pollard model
was the first one that could be fitted to the entire mortality curve, starting
from infancy to very advanced years. We describe it in section \ref{s5}. 
Starting section \ref{s6} we turn our attention to stochastic models, starting 
with the celebrated Lee-Carter model. An interesting class of models used 
across social sciences is the Age-Period-Cohort models. We review a few of them 
relevant to mortality in section \ref{s7}. In section \ref{s8} we consider the 
Cairns-Blake-Dowd model and we consider stochastic models with cohort effects in
 section \ref{s9}. 
The models considered so far have looked at mortality irrespective of the
cause of death. However, in recent years, death rate due to certain causes
is falling while those due to others is rising. One can, therefore, forecast
death rates by cause and aggregate the results. This is the approach taken by
the cause-of-death models in section \ref{s11}. Several mortality models can
be written in the form $f(\mu) = L(\theta)$ where $L$ is a linear function of
the parameters $\theta$ of the model and $f$ is some function of the force
of mortality that permits us to have the right hand side as $L(\theta)$. The
statistically rigorous method to fit such models to mortality data is called
the generalized linear models. Section \ref{s12} discusses them. Not all
models can be brought in this format. Important examples being Lee-Carter and
Renshaw-Haberman. One has to take recourse to spline regression in these
cases. They are the topic of section \ref{s13}. We conclude our review in the
final section.

@@

Gompertz's model and its descendants considered in this section had simple
algebraic forms, each term of which could be interpreted in biological
terms. However, they were not always easy to fit on available mortality
data. The presence of exponential also made them numerically unstable. 
Moreover, they covered only a limited number of sources of mortality.

@@

\section{Generalized linear models}\label{s12}
The ordinary least square linear regression assumes that the residuals are
normally distributed. The expected value of the response is a linear
combination of the predictors. Therefore a constant change in the predictor
results in a proportional change in the response, a characteristic of a linear
relationship. However, often times this is not true. A generalized linear 
model expresses a certain function of the response as a linear model. That is,
instead of fitting a linear model between a response $y$ and the predictors $X$,
we git a linear model between a function $g(y)$ of the response and the
predictors. The function $g$ is called the link function. This method,
developed by Nelder and Wedderburn \cite{nelder1972generalized} proposes a 
general structure for several types of regression including the traditional
ordinary-least-squares, the logistic and the Poisson regressions.

If the probability mass or density function of a random variable $Y$ can be
written as
\begin{equation}\label{s12e1}
f_Y(y, \theta; \phi) = 
\exp\left(\frac{y(\theta) - b(\theta)}{\alpha(\phi)} + c(y, \phi)\right)
\end{equation}
where $a, b, c$ are functions of their arguments and $\theta, \phi$ are
parameters then $Y$ is said to belong to an exponential family of distributions.The normal, binomial, Poisson, exponential and gamma distributions are all
members of the exponential family. A generalized linear model assumes that
the responses $Y_1, \ldots, Y_n$ are independent and identically distributed
with the distribution from the exponential family.

@@

If we write $B = e^{k_1}$ in equation \eqref{s3e3} describing the force of
mortality of Gompertz' law then we can take logarithm of both sides to get
\begin{equation}\label{e12e2}
\log\mu(x) = k_1 + kx.
\end{equation}
The right hand side of this equation is a linear function in $x$ while the left
hand side is a logarithm. This is precisely the form of generalized linear
model with logarithmic link function and a Poisson distribution. If the errors
in Gompertz's law are modeled in terms of a binomial distribution then the
link function is a complimentary log-log function \cite{currie2013fitting} 
defined as
\begin{equation}\label{e12e3}
f(x) = \log(-\log(1 - x)),
\end{equation}
where $0 < x < 1$. Mortality laws expressed as sigmoid functions of the form
\begin{equation}\label{e12e4}
q_x = \frac{e^{\theta_0 + \theta_1 x}}{1 + e^{\theta_0 + \theta_1 x}},
\end{equation}
can be easily transformed to the form
\begin{equation}\label{e12e5}
\frac{q_x}{1 - q_x} = e^{\theta_0 + \theta_1 x}
\end{equation}
or, taking logarithm of both sides,
\begin{equation}\label{e12e6}
\log\frac{q_x}{1 - q_x} = \theta_0 + \theta_1 x.
\end{equation}
Once again we observe that the right hand side is a linear function of the
predictor. The function on the left hand is called a logit function and it is
a legitimate link function for a generalized linear model. In fact, this link
function is used to implement the standard logistic regression in applied
statistics. It is possible to show that the Age-Period-Cohort models and the
Cairns-Blake-Dowd models can also be expressed as generalized linear models. 
The algebra involved to demonstrate this fact is a bit involved and we refer
to the paper by Currie \cite{currie2013fitting} for the details.

@@

Splines were used to draw a smooth curve passing through a set of points. They
were initially used by draftsmen who represented the points by heavy leads and
the splines by flexible wooden strips. This idea was borrowed into numerical
analysis by I. J. Shoenberg \cite{schoenberg1969cardinal} in the late sixties
of the twentieth century. A set of polynomials was used to approximate a
non-linear function where each polynomial fitted a portion of the function
being approximated. It was possible to show that under certain conditions
\cite{schoenberg1969cardinal} a non-linear function could be expressed as a 
Pierce's polynomial function. The polynomials were required to blend into each
other that the boundaries in a smooth way. That is, not only were their values
required to match but also their derivatives.

Splines can be implemented using the Cox-de Boor recurrence relation
\cite{de1978practical}
\begin{equation}\label{s13e1}
B_{i,0}(x) = \begin{cases}
1 & \text{ if } x \in [t_i, t_{i+1}] \\
0 & \text{ otherwise}
\end{cases}
\end{equation}
and
\begin{equation}\label{s13e2}
B_{i,j}(x) = \frac{x - t_i}{t_{i+j} - t_i}B_{i,j-1}(x) + 
\frac{t_{i+j+1}-x}{t_{i+j+1}-t_{i+1}}B_{i+1,j-1}(x).
\end{equation}
These splines are then joined at the knot points $\{t_i: i = 1, \ldots, n\}$.
Once the knots are given it is easy to find the splines. However, there is
little guidance on the number of knots. Too few of them lead to under-fitting
and too many of them lead to over-fitting. O'Sullivan \cite{o1986statistical}
suggested that one should use a relatively large number of knots and prevent
over-fitting by levying a penalty on the second derivatives of the splines.
Eilers and Marx \cite{eilers1996flexible} simplify O'Sullivan's suggestion
and apply it to regression on B-splines. They applied the penalty to the
difference between the coefficients of the adjacent splines. The optimal amount
of smoothing is determined by cross-validation and the Akaike Information 
Criterion (AIC). Eilers and Marx call splines with penalties as P-splines.
They demonstrate that P-splines are an extension of the generalized linear
models. P-spline transforms conserve statistical moments of the data and have
polynomial curve fits as limits. Currie and others \cite{currie2004smoothing}
show how to use P-splines to fit mortality forecasting models. If $x_a = (
x_{a,1},\ldots,x_{a,n_a})^T$ and $x_y=(x_{y,1},\ldots,x_{y,n_y})^T$ are the
vectors with ages and years, $\mathcal{B}_a = \{B_{a,1}, \ldots, B_{a, c_a}\}$
and $\mathcal{B}_y = \{B_{y,1},\ldots,B_{y, c_y}\}$ are the cubic B-splines of
dimensions $c_a$ and $c_y$ then
\begin{equation}\label{c14e1}
X = B_y \otimes B_a
\end{equation}
is the regression matrix for the P-spline model
\begin{equation}\label{c14e2}
M(x,t) = \sum_{i,j}\theta_{i,j}B_{i,j}^{ay}(x, t).
\end{equation}
Here $B_a =(B_{a,j}(x_{a,i}))$ is an $n_a \times c_a$ matrix and $B_y = 
(B_{y,j}(x_{y,i}))$ is an $n_y \times c_y$ matrix. The penalty matrix is given
as
\begin{equation}\label{c14e3}
P = \lambda_aI_{c_y} \otimes D_a^TD_a + \lambda_y D_y^TD_y\otimes I_{c_a},
\end{equation}
where $D_a$ is the difference matrix of the rows of M, $D_y$ is the difference
matrix on the columns of M, and the two $\lambda$s are the smoothing parameters.
The regression problem is then minimizing the likelihood function
\begin{equation}\label{c14e4}
\ell_p = \ell - \frac{1}{2} m^TPm,
\end{equation}
where $\ell$ is the default log likelihood function and $m$ is the vector of
the elements of $M$.
